Generating Empirically Optimized Composed Matrix Kernels from MATLAB Prototypes

### 2009-02-13 doc/papers/iccs09/paper.tex
Final changes to address reviewer comments (mainly Reviewer 3)
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@441 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

To generate highly optimized code from a MATLAB prototype of the composed
BLAS operation, we follow a three-step approach, illustrated in
Figure~\ref{fig:process}.  
%
\begin{wrapfigure}{r}{3in}
\vspace{-.2in}
\centering
\includegraphics[width=3in]{figures/process.png}
\caption{Code generation and tuning process.}
\label{fig:process}
\vspace{-.2in}
\end{wrapfigure}
%
To begin, we have developed a compiler that
converts a MATLAB script into simple C code~\cite{Siek}.  After generating
the C code from the high-level MATLAB prototype, we (optionally) use 
the source-to-source automatic parallelization tool Pluto~\cite{Pluto} %developed at Ohio State University
to optimize for coarse-grained parallelism and locality simultaneously. Using
the results of the Pluto analysis, we insert annotations into the C code,
which are then processed by our extensible annotation system Orio to generate
many tuned versions of the same operation using different optimization
parameters. Orio then performs an empirical search to select the best among
the multiple optimized code variants.

@@

Figure~\ref{fig:compiler} gives an overview of the MATLAB-to-C compilation process~\cite{Siek}. 
The MATLAB kernel specification is parsed into a high-level intermediate
representation in the form of a dataflow graph, in which each node represents
a parameter (e.g., a scalar, matrix, or vector variable) of the kernel or an operation.
%
\begin{wrapfigure}{r}{2in}
%\vspace{-.2in}
\centering
\includegraphics[width=2in]{figures/compile.png}
\caption{MATLAB-to-C compiler.}
\label{fig:compiler}
\vspace{-.2in}
\end{wrapfigure}
%
This dataflow graph is then iteratively processed until all of
the implementation choices have been made.  The compilation process consists
of three phases -- analysis, refinement, and optimization -- that are together
iterated until all of the implementation decisions have been made.  The graph
is then translated into C code.

@@

Figure~\ref{fig:orio} illustrates the tuning process implemented in
Orio. The input to Orio is C code containing semantic comments that
describe both the computation and various performance-tuning
directives. Orio first extracts all annotation code regions by parsing
the marked-up input code. Each annotated region is then passed to code
transformation module and code generator for potential optimizations. Next the
transformed C code with various incorporated optimizations
corresponding to the specified annotations is produced. The
source-to-source transformation system of Orio generates an optimized
code version for each distinct combination of performance parameter
values. Each generated code variant is then executed and its
performance cost evaluated. After iteratively testing all code
variants, the best-performing code is selected as the final output of
Orio. Because the search space of all possible optimized code variants
can be exponentially large, the search engine implements a number of
search heuristics (i.e., random, simplex, and simulated annealing) to
effectively narrow the search for near-optimal performance. The tuning
specifications, written by users in the form of annotations, contains
information required for guiding the transformation and tuning
process.

\begin{figure}[tb]
%\vspace{-.3in}
\begin{center}
\includegraphics[width=.65\textwidth]{figures/orio.png}
\end{center}
\vspace{-.2in}
\caption{Overview of the Orio empirical tuning process.}
\label{fig:orio}
\vspace{-.2in}
\end{figure}

@@

Figure~\ref{fig:orio-example} shows an annotation example used by Orio to
empirically optimize VADD operation on Blue Gene/P. The annotations contain
performance hints that instruct Orio to perform memory alignment
optimization, loop unrolling, and multicore parallelization (using
OpenMP). In addition to these simple optimizations, Orio supports other
transformations such as loop blocking, loop permutation, scalar replacement,
array copy optimization, and some architecture-dependent optimizations. The
right-hand side of Figure~\ref{fig:orio-example} shows separate tuning
specifications used for building and running executable tests, including
performance parameter values, execution environment details, input variable
information, and the search algorithm. Orio also supports parallel search
when parallel resources are available. In this example, the parallel Orio
driver simultaneously executes 64 code variants in the same parallel job. The
commands used by Orio to submit a parallel job and to query its status are
also specified in the tuning specifications. At present, users must create
the tuning specifications manually. When Orio is used in conjunction with
compiler tools, such as the MATLAB compiler described in this paper, it
should eventually be possible to automatically generate the tuning
specifications.

@@

\begin{minipage}[b]{.5\textwidth}
\includegraphics[width=\textwidth]{figures/vadd_bgp.png}
\end{minipage}
&
\begin{minipage}[b]{.5\textwidth}
\includegraphics[width=\textwidth]{figures/atax.png}
\end{minipage}\\

@@

\begin{minipage}[b]{.5\textwidth}
\includegraphics[width=\textwidth]{figures/gemver.png}
\end{minipage}
&
\begin{minipage}[b]{.5\textwidth}
\includegraphics[width=\textwidth]{figures/bicgkernel.png}
\end{minipage}\\

@@

The performance results (in MFLOP/s) of tuning the VADD operation on the Blue
Gene/P are given in Figure~\ref{fig:results}(a). The ``Base'' label
designates the C implementation generated by the MATLAB compiler. We also
tested the performance of an implementation that calls
\texttt{DAXPY} twice using available BLAS libraries.
Finally, we tuned the simple C loop version using Orio, with the performance
annotations previously shown in Figure~\ref{fig:orio-example}. In this
experiment, we measured the performance for both the sequential and parallel
scenarios. Even for a very simple operation such as vector addition the
compiler alone is unable to obtain the same level of performance as the
empirically tuned versions. Furthermore, as expected, the BLAS implementation
does not exploit locality and thus performed worse than the single-loop
implementation.
%of the single loop implementation due to l
% however, selecting the correct compiler
%options was crucial for achieving good performance of the base version.

The experiments of the remaining operations were performed on the
multicore Intel Xeon. Included in these experiments are performance
numbers for six code variants: the C code generated by the MATLAB
compiler (``C from MATLAB''), three BLAS-based implementations that
use Intel MKL, ATLAS, and the default BLAS library on Ubuntu 8.04, and
the sequential and parallel code variants tuned by Orio (``Orio
(Seq.)'' and ``Orio (Par.)'', respectively).

>>>>

To generate highly optimized code from a MATLAB prototype of the composed
BLAS operation, we follow a three-step approach, illustrated in
Figure~\ref{fig:process}.  This process is repeated when the MATLAB code changes
or the code must be tuned on a new architecture.
To begin, we have developed a compiler that
%
\begin{wrapfigure}{r}{3in}
\vspace{-.2in}
\centering
\includegraphics[width=3in]{figures/process.pdf}
\vspace{-.15in}
\caption{Code generation and tuning process.}
\label{fig:process}
\vspace{-.1in}
\end{wrapfigure}
%
converts a MATLAB script into simple C code~\cite{Siek}.  After generating
the C code from the high-level MATLAB prototype, we (optionally) use 
the source-to-source automatic parallelization tool Pluto~\cite{Pluto} %developed at Ohio State University
to optimize for coarse-grained parallelism and locality simultaneously. Using
the results of the Pluto analysis, we insert annotations into the C code,
which are then processed by our extensible annotation system Orio to generate
many tuned versions of the same operation using different optimization
parameters. Orio then performs an empirical search to select the best among
the multiple optimized code variants.

@@

Figure~\ref{fig:compiler} gives an overview of the MATLAB-to-C compilation process~\cite{Siek}. 
The MATLAB kernel specification is parsed into a high-level intermediate
representation in the form of a dataflow graph, in which each node represents
a parameter (e.g., a scalar, matrix, or vector variable) of the kernel or an operation.
%
\begin{wrapfigure}{r}{2in}
%\vspace{-.2in}
\centering
\includegraphics[width=2in]{figures/compile.pdf}
\caption{MATLAB-to-C compiler.}
\label{fig:compiler}
\vspace{-.2in}
\end{wrapfigure}
%
This dataflow graph is then iteratively processed until all of
the implementation choices have been made.  The compilation process consists
of three phases -- analysis, refinement, and optimization -- that are together
iterated until all of the implementation decisions have been made.  The graph
is then translated into C code.

@@

Figure~\ref{fig:orio} illustrates the tuning process implemented in
Orio. The input to Orio is C code containing semantic comments that
describe both the computation (using a syntax that is more restricted
than the original C) and various performance-tuning directives. Orio
first extracts all annotation code regions by parsing the marked-up
input code. Each annotated region is then passed to code
transformation module and code generator for potential
optimizations. Next the transformed C code with various incorporated
optimizations corresponding to the specified annotations is
produced. Orio generates an optimized code version for each distinct
combination of performance parameter values. Each code
variant is then executed and its performance measured. After
iteratively testing all code variants, the best-performing code is
selected as the final output of Orio. While each variant is
computationally very cheap (a single code variant takes between a
fraction of a second to a few seconds depending on the input sizes),
the search space of all possible optimized code variants can be
exponentially large. Therefore, the search engine implements a number
of search heuristics (i.e., random, simplex, and simulated annealing)
to effectively narrow the search for near-optimal performance and
reduce the empirical search time. 
%The tuning
%specifications, written by users in the form of annotations, contains
%information required for guiding the transformation and tuning
%process.

\begin{figure}[tb]
%\vspace{-.3in}
\begin{center}
\includegraphics[width=.65\textwidth]{figures/orio.pdf}
\end{center}
\vspace{-.2in}
\caption{Overview of the Orio empirical tuning process.}
\label{fig:orio}
\vspace{-.15in}
\end{figure}

@@

Figure~\ref{fig:orio-example} shows an annotation example used by Orio to
empirically optimize VADD operation on Blue Gene/P. The annotations contain
performance hints that instruct Orio to perform memory alignment
optimization, loop unrolling, and multicore parallelization (using
OpenMP). In addition to these simple optimizations, Orio supports other
transformations such as loop blocking, loop permutation, scalar replacement,
array copy optimization, and some architecture-dependent optimizations. The
right-hand side of Figure~\ref{fig:orio-example} shows separate tuning
specifications used for building and running executable tests, including
performance parameter values, execution environment details, input variable
information, and the search algorithm. Orio also supports parallel search
when parallel resources are available. In this example, the parallel Orio
driver simultaneously executes 64 code variants in the same parallel job. 
%The
%commands used by Orio to submit a parallel job and to query its status are
%also specified in the tuning specifications. 
At present, users must create the tuning specifications manually for
each architecture. When Orio is used in conjunction with compiler
tools, such as the MATLAB compiler described in this paper, it should
eventually be possible to automatically generate the tuning
specifications.

@@

\begin{minipage}[b]{.5\textwidth}
\includegraphics[width=\textwidth]{figures/vadd_bgp_time2.pdf}
\end{minipage}
&
\begin{minipage}[b]{.5\textwidth}
\includegraphics[width=\textwidth]{figures/atax.pdf}
\end{minipage}\\

@@

\begin{minipage}[b]{.5\textwidth}
\includegraphics[width=\textwidth]{figures/gemver.pdf}
\end{minipage}
&
\begin{minipage}[b]{.5\textwidth}
\includegraphics[width=\textwidth]{figures/bicgkernel.pdf}
\end{minipage}\\

@@

The performance results (in MFLOP/s) of tuning the VADD operation on the Blue
Gene/P are given in Figure~\ref{fig:results}(a). The ``Base'' label
designates the C implementation generated by the MATLAB compiler. We also
tested the performance of an implementation that calls
\texttt{DAXPY} twice using available BLAS libraries.
Finally, we tuned the simple C loop version using Orio, with the performance
annotations previously shown in Figure~\ref{fig:orio-example}. In this
experiment, we measured the performance for both the sequential and parallel
scenarios (indicated by (S) and (P) in the legend, respectively). Even for a very simple operation such as vector addition the
compiler alone is unable to obtain the same level of performance as the
empirically tuned versions. Furthermore, as expected, the BLAS implementation
does not exploit locality and thus performed worse than the single-loop
implementation.
%of the single loop implementation due to l
% however, selecting the correct compiler
%options was crucial for achieving good performance of the base version.

The experiments of the remaining operations were performed on the
multicore Intel Xeon. Included in these experiments are performance
numbers for six code variants: the C code generated by the MATLAB
compiler (``C from MATLAB''), three BLAS-based implementations that
use Intel MKL, ATLAS, and the default BLAS library on Ubuntu 8.04, and
the sequential and parallel code variants tuned by Orio (``Orio
(S)'' and ``Orio (P)'', respectively).
